### 2012-06-26 trunk/ijcis-pos-model/sources/temporaldatabases.tex
final changes

\begin{definition}
\emph{Valid-time relation.}
Consider $\A$ be the set of all the entities, and let $A=\left(A_1, \ldots, A_n \right), A \in \A$ be the set of attributes that define an entity. Let $I = \left(S_i, E_i \right)$ be the set of attributes defining the time interval. The set $\I$ is the set of all the time intervals. The pair $\left(A, I\right)$ expresses that the data regarding the entity A are valid during the crisp time interval I. Let $R \subseteq \A \  x\  \I$ be a valid-time relation. Then, the following equation indicates that the pair $\left(A, I\right)$ is in the relation R:

\begin{equation}
\label{eq:rel-def}
\left( R, \left(A , I\right) \right)
\end{equation}
\end{definition}

In order to simplify the algorithms for the manipulation of data, some auxiliary functions and constants are defined:

\begin{definition}
 \label{def:from-beginning}
\emph{From the beginning (FB).} 
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $I = \left(S_i, E_i \right), S_i, E_i \in \T$ the set of attributes defining the time interval. Let $R \subseteq \A\  x\  \I$ be a valid-time relation. We will say $I = \left(FB, E_i \right)$ when $I = \left(- \infty, E_i \right)$.
%The from the beginning constant is used when the value for $s_i$ is equal to $-\infty$.
\end{definition}

\begin{definition}
 \label{def:until-changed}
\emph{Until changed (UC).} 
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity. Let $\T$ be the time domain, and $I = \left(S_i, E_i \right), S_i, E_i \in \T$ the set of attributes defining the time interval. Let $R \subseteq \A\  x\  \I$ be a valid-time relation. We will say  $I = \left(S_i, UC \right)$ when $I = \left(S_i, +\infty \right)$.
%The until changed constant is used when the value for $e_i$ is equal to $+\infty$.
\end{definition}

If the pair $\left(A, I \right)$ with $I = \left(S_i, UC \right), S_i \in \T$ is in the relation $R$, then, it is said that the pair $\left(A, I \right)$ is \emph{current} in the relation.

For example, let us consider the last row in Table \ref{tbl:library-sample}. The value for the time interval is $I = \left(4/4/2012 , UC \right)$. The meaning is that the patient with ID = 3 was in the hospital the  4/4/2012 and he or she is still there. The patient with ID = 3 is current in the relation.

\begin{example}
\label{ex:library-database}
 Consider a hospital database. Each patient is stored in the database with an identifier (ID), the name and the dates when the patient entered and left the hospital. Table \ref{tbl:library-sample} illustrates the first version of the database, after three insertions. In this example, $A = ($ ID, Name $)$ and $I = ($ Arrival, Leave $)$.
\end{example}

\begin{samepage}
\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample}Sample hospital database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Name  & Arrival & Leave \\
\hline
3 & John & 15/3/2012 & 30/3/2012 \\
4 & Antoon & 25/3/2012 & 4/4/2012 \\
5 & Daan & 18/3/2012 & 2/4/2012\\
3 & John & 4/4/2012 & UC \\
\hline\\
\end{tabular}
}
\end{samepage}

\begin{definition}
\label{def:current-in-relation}
\emph{Current $\left(R, A \right)$.}
Consider $A=\left(A_1, \ldots, A_n \right), A \in \A$ be the set of attributes that define an entity and $I = \left(S_i, UC \right)$ the set of attribute defining the time interval. Let $R \subseteq \A\  x\  \I$ be a valid-time relation. The function Current $\left(R, A \right)$ returns the crisp time interval $I$ if the entity $A$ is currently in the relation $R$ or the empty set, if the entity $A$ is not currently in the relation. It is defined as follows:

@@

\begin{align}
\label{eq:current-in-relation}
\mbox{Current} \left(R, A \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } (A,I) \in R \mbox{ and } I = \left(S_i, UC \right), S_i \in \T\\
\emptyset & \mbox{ otherwise}
\end{cases}
\end{align}
\end{definition}

For example, the function Current(R, (3,`John')) returns the time interval $I = \left(4/4/2012 , UC  \right)$. Conversely, Current(R, (4,`Antoon')) returns the empty set.

@@

\begin{definition}
\emph{Overlaps:}
 \label{def:overlaps}
Given two time intervals $I = \left(S_i, E_i \right)$ and $J = \left(S_j, E_j\right)$, it is said that I overlaps J if:
\begin{equation}
 \label{eqn:overlaps}
\text{I Overlaps J}  = \left(S_i > S_j  \right) \wedge \left(E_i < E_j  \right) 
\end{equation}
\end{definition}

\begin{definition}
\emph{During:}
 \label{def:during}
Given two time intervals $I = \left(S_i, E_i \right)$ and $J = \left(S_j, E_j\right)$, it is said that I during J if:
\begin{equation}
 \label{eqn:during}
\text{I During J}  = \left(S_i < S_j  \right) \wedge \left(E_i > S_j  \right) \wedge \left(E_i < E_j \right)
\end{equation}
\end{definition}

@@

\begin{definition}
\emph{CloseR$\left( I, J \right)$:}
\label{def:close-a-crisp-interval-r}
Consider two crisp intervals $I= \left[S_i ,E_i \right]$ and $J= \left[S_j ,E_j \right]$. The CloseR$\left(I, J\right)$ function allows to close the right-open interval $I$ with respect to the first value $e$ in $J$:
\begin{align}
\label{eq:close-a-crisp-interval}
\mbox{CloseR} \left( I, J \right) &=& \\ 
\begin{cases}
\nonumber
%I & \mbox{ if } b \neq +\infty \\
%I=\left[a, c \right[ & \mbox{ if } b = +\infty \wedge J > I
\left[S_i, S_j \right[ & \mbox{ if } E_i = UC \mbox{ and } J \mbox{ During } I \\
I & \mbox{ otherwise}
\end{cases}
\end{align}
\end{definition}

For example, consider two time intervals, $I = \left(4/4/2012, UC \right)$ and $J = \left(24/4/2012, UC \right)$. The result of applying the CloseR$\left(I, J \right)$ is $I = \left(4/4/2012, 23/4/2012 \right)$.

@@

\begin{definition}
\label{def:close-current-version}
\emph{Close-current$\left(R, \left(A, J  \right)\right)$}.
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $J = \left(S_j,E_j \right)$ the set of attribute defining the time interval. Let $R \subseteq \A\  x\  \I$ be a valid-time relation. Then, the function close-current$\left(R,\left(A,J\right) \right)$ closes any current version of the object $A$ if it exists and add the new version $\left(A, J\right)$.

\begin{eqnarray}
\label{eq:close-current}
\text{Close-current} \left(R, \left(A, J\right) \right) =\\
\begin{cases}
\nonumber
R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseR } \left(I, J\right) \right)\right \rbrace \cup \left \lbrace\left(A, J\right) \right \rbrace  \\
\nonumber
\mbox{ if  Current } \left(R, A \right) \neq \emptyset\\ %\wedge J > \mbox{ CloseR } \left(I, J \right)   \\
\nonumber R , \text{ otherwise}
\end{cases}
\end{eqnarray}
\end{definition}

For example, consider the patient with ID = 3 in Table \ref{tbl:library-sample}. The current version arrived on 4/4/2012. The patient left the hospital but, for some reason, the date  was not registered. The patient arrived again at the hospital on 24/4/2012. Then, the Close-current function allows to close the current version of the patient and insert a new version. First, the function CloseR is applied with $I = \left(4/4/2012, UC \right)$ and $J = \left(24/4/2012, UC \right)$. Hence, the value for the time interval $I$ is $ \left(4/4/2012, 23/4/2012 \right)$. Then, the modifications on the value for the time interval $I$ are stored. Finally, a new row with the current values of the patient and the time interval $J$ is stored.
The result of this operation is illustrated on Table \ref{tbl:library-sample-2}.

%It is also possible to append at the beggining by using \eqref{def:close-a-crisp-interval-l} and \eqref{def:first-in-relation}.
%\begin{definition}
%\label{def:append-first-version}
%Consider an object $A$, the relation $R$ and a time interval $J$. Then, the function append-first$\left(R,A,J \right)$ closes any right open version of the object $A$ if it exists and add the new version $\left(A, J \right)$.
%
%\begin{eqnarray}
%%\label{eq:close-current}
%\text{Append-first} \left(R, A, J \right) =\\
%\begin{cases}
%\nonumber
%\big \lbrace R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseL } \left(I, J\right) \right) \cup \left(A, J\right)\right \rbrace  \big \rbrace \\
%\nonumber
%\mbox{ if } I = \mbox{ First } \left(R, A \right) \wedge J < \mbox{ CloseL } \left(I, J \right)   \\
%\nonumber R , \text{ in any other case}
%\end{cases}
%\end{eqnarray}
%\end{definition}
\subsubsection{\label{subsubsec:modify}Modify}
This operation adds new information about an existing entity $A$ in the relation $R$. The modify operation does not remove any previous value of the object $A$. It closes the current version for the entity $A$ and adds a new version.

@@

\begin{definition}
 \emph{Modify(R, (A,I))}.
Let $R \subseteq \A\  x\  \I$ be a valid-time relation and  $A=\left(A_1, \ldots, A_n \right), A \in \A$ be the set of attributes that define an entity and $I = \left(S_i,E_i\right)$ the set of attributes defining the time interval. The algorithm for the modify operation is defined as follows.
\end{definition}

\begin{align}
\label{eq:modify}
\MOD \left( R, \left(A, I \right)\right) =\\
%\begin{cases}
\nonumber
\mbox{ Close-current }\left(R, \left(A, I\right) \right) %\\
% \mbox{ if Current}\left(R ,A \right) \neq \emptyset \\
% R,  \mbox{ otherwise }
%\end{cases}
\end{align}

\subsubsection{\label{subsubsec:insert}Insert}
The user wants to store an entity $A$ which is valid in the relation $R$ during the time interval specified by $I = \left(a, b \right)$.
%
%
%The interpretation of the insert operation is the following: The user wants to store in the database that an object $A$  is true for some period(s) of time $I$ in the given relation $R$. To indicate that the object $A$ is current in the relation, the special value Until Current, $\UC$  is used. 
There are two main cases when performing a create operation:
\begin{enumerate}
\item The entity $A$ was never in the relation $R$: The entity is added with the valid-time indicated by the crisp interval $I$. 
% In the example, this correspond with the following insertion sentences:
% \begin{verbatim}
% Insert(3,'Dracula',
% (15/3/2012, 30/3/2012));
% Insert(4,'Frankenstein',
% (25/3/2012, 4/4/2012));
% Insert(5,'Harry Potter',
% (18/3/2012, 2/4/2012));
% \end{verbatim}

\item The entity $A$ is in the relation $R$. Depending on the value of the time interval $I$, there are three possibilities:
	\begin{enumerate}
	\item Insert $\left(A, I\right)$ in the relation $R$. If the time interval $I$ does not overlap any other valid-time interval in the relation $R$ for the entity $A$. For example, consider that the book with ID = 3 is loaned on 4/4/2012 but the book has not been returned yet. This is illustrated in Table \ref{tbl:library-sample}.
%  The insert sentence is:
% 	      \begin{verbatim}
% Insert(3,'Dracula',
% (4/4/2012, UC));
% 	      \end{verbatim}

	\item Modify and close the current version of $A$ and insert $\left(A, I\right)$. For example, consider now that the patient with ID = 3 arrived to the hospital on 24/4/2012. Here the problem is that the patient with ID = 3 arrived on 4/4/2012 but, for some reason, the departure date was not stored. If the patient is already at the hospital, then it is necessary to set the departure date and add a new row with the new arrival date. This is illustrated in Table \ref{tbl:library-sample-2}.
%  the insert sentence is:
% 	    \begin{verbatim}
% Insert(3,'Dracula',
% (24/4/2012, UC));
% 	      \end{verbatim}
	\item Reject the insertion, if the time interval $I$ does overlap any existing valid-time interval for the entity $A$ in the relation $R$. For example, consider that the hospital manager wants to introduce a past arrival for the patient with ID = 3. The patient arrived on 6/4/2012 and left the hospital on 25/4/2012. As the dates do overlap, it is not possible that the patient was in the hospital at that time interval. Therefore, the insertion is rejected.
%  The insert sentence is:
% 
% 	      \begin{verbatim}
% Insert(3,'Dracula',
% (6/4/12/2012,25/4/2012));
% 	    \end{verbatim}

@@

\end{enumerate}
\begin{samepage}
\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample-2}Sample hospital database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Name  & Arrival & Leave \\
\hline
3 & John & 15/3/2012 & 30/3/2012 \\
4 & Antoon & 25/3/2012 & 4/4/2012 \\
5 & Daan Potter & 18/3/2012 & 2/4/2012\\
3 & John & 4/4/2012 & 23/4/2012 \\
3 & John & 24/4/2012 & UC \\
\hline\\
\end{tabular}
}
\end{samepage}

\begin{definition}
 \emph{Insert(R,(A,I)))}.
Let $R \subseteq \A\  x\  \I$ be a valid-time relation and  $A=\left(A_1, \ldots, A_n \right), A \in \A$ be the set of attributes that define an entity and $I = \left(S_i,E_i\right)$ the set of attributes defining the time interval. Then, the algorithm for the implementation of the insert operation is defined as follows.
\end{definition}

>>>>

\begin{definition}
\label{def:valid-time-relation}
\emph{Valid-time relation.}
Consider the following definitions and notations:

\begin{itemize}
 \item A set of non-temporal attributes.
	\begin{equation}
	\label{eq:attribute-set}
	A = \left \lbrace A_1, A_2, \ldots, A_n \right \rbrace
	\end{equation}
      The domain for each attribute $A_1, \ldots, A_n$ is $D_1, \ldots, D_n$ respectively. 
\item The original primary key $A_K$ is a subset of the attributes in $A$.
      \begin{equation}
       \label{eq:primary-key-a}
      A_K \subset A
      \end{equation}
\item A set of two attributes, $S$  and $E$ for the starting and ending points respectively. $I$ defines the valid time interval for the data. 
\begin{equation}
 \label{eq:attribute-time-interval}
I = \left \lbrace S, E \right \rbrace
\end{equation}
\item Then $R$, the schema for the valid-time relation is:
\begin{equation}
 \label{eq:valid-time-relation}
R = A \cup  I
\end{equation}
\item The primary key for the valid-time relation $R$ is:
\begin{equation}
 \label{eq:valid-time-temporal-pk}
PK = A_K \cup I
\end{equation}

\item We will note by $r$ any valid instance of $R$. 
      \begin{equation}
       \label{eq:valid-time-instance}
      r \subseteq D_1\ x\ \ldots\ x\ D_n
      \end{equation}

\item Let $t$ be a tuple in the instance $r$; $t \in r$.
    % \begin{itemize}
    %  \item
Then, we will note the values for the starting and the ending  points, $s$ and $e$  respectively. The value for the time interval is given by $i$.
      \begin{align}
       \label{eq:starting-point}
      s = t\left[S \right]\\
      e = t\left[E \right]\\
      i = \left(s, e\right)
      \end{align}

      \item Let $A_k \subset A$ be as defined in \eqref{eq:primary-key-a}). Then, $a_k$ denotes the values for $A_K$ in tuple $t$.
	    \begin{equation}
	     \label{eq:pk-attribute} 
	      a_k = t\left[A_K \right]
	    \end{equation}

    \item Let $PK$ be the primary key for the valid-time relation as given in equation \eqref{eq:valid-time-temporal-pk}. Then, $pk$ denotes the values for the attributes in the primary key in a tuple $t$.
	  \begin{equation}
	   \label{eq:value-pk}
	  pk = t\left[PK \right]
	  \end{equation}

     % \end{itemize}
\item $Vt$ is the set of all the versions for a given tuple $t$. Formally,

\begin{equation}
 \label{eq:all-the-versions}
Vt =\left \lbrace t_i \in r, t_i\left[A_K\right] = t\left[A_K\right] \right \rbrace
\end{equation}
Obviously, $t$ itself is included in the set.
% If a tuple $t$ contains more than one version, the result of $Vt$ is a set. We will use the index $k$ to address the elements of the set. E.g. $Vt = \left \lbrace t_1, t_2, \ldots, t_n \right \rbrace$. Then $t_k , k \in \left \lbrace 1, \ldots, n \right \rbrace$ is an element of $Vt$.
\end{itemize}
\end{definition}
We will illustrate the definitions with an example.

\begin{example}
Consider the set of attributes $A = \left \lbrace A_1, A_2, A_3 \right \rbrace$. The primary key for these attributes is given by $A_K = \left \lbrace A_1, A_2 \right \rbrace$. Let $I = \left \lbrace S, E \right \rbrace$ be the set of temporal attributes that define the validity period of the data. $R$ is the valid-time relation and $r$ is an instance of the relation. The instance $r$ is given by the following elements. $r = \left \lbrace \left(a_{11}, a_{12}, a_{13}, s_1 ,e_1 \right) \right.$,  $\left(a_{21}, a_{22}, a_{23}, s_2, e_2 \right)$ , $\left. \left(a_{11}, a_{12}, a_{31}, s_3, e_3 \right) \right \rbrace$. The instance $r$ is illustrated in Table \ref{tbl:sample-definitions}. 
Consider the tuple $t = \left(a_{11}, a_{12}, a_{13}, s_1 ,e_1 \right)$. Then,
\begin{align}
 \nonumber
s_1 &= t\left[S \right]\\
 \nonumber
e_1 &= t[E]\\
 \nonumber
i\ \ &= \left(s_1, e_1\right)\\
 \nonumber
a_k &= t\left[A_K\right] = \left(a_{11}, a_{12} \right) \\
 \nonumber
pk &= t\left[PK\right] =\left(a_{11}, a_{12}, s_1, e_1\right)\\
 \nonumber
Vt &= \left \lbrace t_1, t_3 \right \rbrace
\end{align}

\end{example}

\begin{samepage}
\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:sample-definitions}Sample database containing the instance $r$ of the valid-time relation $R$.}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c c c}\\
\hline
& \textbf{A$_1$}  & \textbf{A$_2$}  & $A_3$ & $S$ & $E$ \\
\hline
$t_1$&$a_{11}$ & $a_{12}$ & $a_{13}$ & $s_1$ & $e_1$ \\
$t_2$ & $a_{21}$ & $a_{22}$ & $a_{23}$ &  $s_2$ & $e_2$ \\
$t_3$ & $a_{11}$ & $a_{12}$ & $a_{31}$ & $s_3$ & $e_3$\\
\hline\\
\end{tabular}
}
\end{samepage}

% Consider $\A$ be the set of all the entities, and let $A=\left(A_1, \ldots, A_n \right), A \in \A$ be the set of attributes that define an entity. Let $I = \left(S_i, E_i \right)$ be the set of attributes defining the time interval. The set $\I$ is the set of all the time intervals. The pair $\left(A, I\right)$ expresses that the data regarding the entity A are valid during the crisp time interval I. Let $R \subseteq \A \  x\  \I$ be a valid-time relation. Then, the following equation indicates that the pair $\left(A, I\right)$ is in the relation R:
% 
% \begin{equation}
% \label{eq:rel-def}
% \left( R, \left(A , I\right) \right)
% \end{equation}
% 
% 
In order to simplify the algorithms for the manipulation of data, some auxiliary functions and constants are defined:

\begin{definition}
 \label{def:from-beginning}
\emph{From the beginning (FB).} 
Consider the elements in definition \ref{def:valid-time-relation}. We will say $i = \left(FB, e \right)$ when $i = \left(- \infty,e \right)$. 
% $i = \left(s, e \right)$ be the value for a time interval as given equation \eqref{}.
% $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $I = \left(S_i, E_i \right), S_i, E_i \in \T$ the set of attributes defining the time interval. Let $R \subseteq \A\  x\  \I$ be a valid-time relation. We will say $I = \left(FB, E_i \right)$ when $I = \left(- \infty, E_i \right)$.
%The from the beginning constant is used when the value for $s_i$ is equal to $-\infty$.
\end{definition}

\begin{definition}
 \label{def:until-changed}
\emph{Until changed (UC).} 
Consider the elements in definition \ref{def:valid-time-relation}. We will say  $i = \left(s, UC \right)$ when $i = \left(s, +\infty \right)$.
% Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity. Let $\T$ be the time domain, and $I = \left(S_i, E_i \right), S_i, E_i \in \T$ the set of attributes defining the time interval. Let $R \subseteq \A\  x\  \I$ be a valid-time relation. We will say  $I = \left(S_i, UC \right)$ when $I = \left(S_i, +\infty \right)$.
%The until changed constant is used when the value for $e_i$ is equal to $+\infty$.
\end{definition}

\begin{definition}
\emph{Current.} 
Consider the elements in definition \ref{def:valid-time-relation}. We will say that the tuple $t$ is current in the instance $r$ of the relation $R$ when $i = \left(s, UC \right)$.
\end{definition}

%If the pair $\left(A, I \right)$ with $I = \left(S_i, UC \right), S_i \in \T$ is in the relation $R$, then, it is said that the pair $\left(A, I \right)$ is \emph{current} in the relation.

For example, let us consider the last row in Table \ref{tbl:library-sample}. The value for the time interval is $I = \left(4/4/2012 , UC \right)$. The meaning is that the document with ID = 3 was valid the  4/4/2012 and it is still valid. The document with ID = 3 is current in the relation.

\begin{example}
\label{ex:library-database}
 Consider a historical database containing diplomatic documents. The starting and the ending dates when the diplomatic document is valid. It is possible that a diplomatic document is valid for a period of time and several years later it becomes valid again. The following elements are stored: an identifier of the document (ID), the entity that issues the document and the dates when document is valid. Table \ref{tbl:library-sample} illustrates the first version of the database, after three insertions. In this example, $A = ($ ID, Entity $)$ and $I = ($ Start, End $)$.
\end{example}

\begin{samepage}
\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample}Sample historical database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Entity  & Start & End \\
\hline
3 & E.U. & 15/3/2012 & 30/3/2012 \\
4 & N.A.T.O.  & 25/3/2012 & 4/4/2012 \\
5 & C.E.I. & 18/3/2012 & 2/4/2012\\
3 & E.U. & 4/4/2012 & UC \\
\hline\\
\end{tabular}
}
\end{samepage}

\begin{definition}
\label{def:current-in-relation}
\emph{Current $\left(r,a_k \right)$.}
Consider the elements in definition \ref{def:valid-time-relation}. The function Current $\left(r,a_k \right)$ returns the crisp time interval $i$ if the tuple $t$ with the primary key values given by $a_k$ is currently in the instance $r$ of the relation $R$. The function returns the empty set otherwise.
% 
% Consider $A=\left(A_1, \ldots, A_n \right), A \in \A$ be the set of attributes that define an entity and $I = \left(S_i, UC \right)$ the set of attribute defining the time interval. Let $R \subseteq \A\  x\  \I$ be a valid-time relation. The function Current $\left(R, A \right)$ returns the crisp time interval $I$ if the entity $A$ is currently in the relation $R$ or the empty set, if the entity $A$ is not currently in the relation. It is defined as follows:

@@

\begin{align}
\label{eq:current-in-relation}
\mbox{Current} \left(r, a_k\right) &=& \\ 
\begin{cases}
\nonumber
t & \mbox{ if } \exists t \in r : i = \left(s, UC \right) \mbox{ and }\\
& a_k = t[A_K] \\
\emptyset & \mbox{ otherwise}
\end{cases}
\end{align}
\end{definition}

For example, the function Current(r, (3,`E.U.')) returns the time interval $i = \left(4/4/2012 , UC  \right)$. Conversely, Current(r, (4,`N.A.T.O.')) returns the empty set.

@@

\begin{definition}
\emph{Overlaps:}
 \label{def:overlaps}
Given two time intervals $i_1 = \left(s_1, e_1 \right)$ and $i_2 = \left(s_2 e_2\right)$, it is said that $i_1$ overlaps $i_2$ if:
\begin{equation}
 \label{eqn:overlaps}
i_1 \text{ Overlaps } i_2  = \left(s_1 < s_2  \right) \wedge \left(e_1 < e_2  \right) 
\end{equation}
\end{definition}

\begin{definition}
\emph{During:}
 \label{def:during}
Given two time intervals $i_1 = \left(s_1, e_1 \right)$ and $i_2 = \left(s_2 e_2\right)$, it is said that $i_1$ during $i_2$ if:
\begin{align}
 \label{eqn:during}
i_1\text{ During } i_2  =& \left(s_1 > s_2  \right) \wedge \left(e_1 > s_2 \right) \wedge \\
\nonumber & \left(e_1 < e_2 \right)
\end{align}
\end{definition}

@@

\begin{definition}
\emph{CloseR$\left(i_1, i_2\right)$:}
\label{def:close-a-crisp-interval-r}
Consider two crisp intervals $i_1 = \left(s_1, e_1 \right)$ and $i_2 = \left(s_2, e_2\right)$. The CloseR$\left(i_1, i_2 \right)$ function allows to close the right-open interval $i_1$ with respect to the first value $s_2$ in $i_2$:
\begin{align}
\label{eq:close-a-crisp-interval}
\mbox{CloseR} \left[ i_1, i_2\right[ &=& \\ 
\begin{cases}
\nonumber
%I & \mbox{ if } b \neq +\infty \\
%I=\left[a, c \right[ & \mbox{ if } b = +\infty \wedge J > I
i_1 = \left(s_1,s_2 \right( & \mbox{ if } e_1 = UC \mbox{ and } i_2 \mbox{ During } i_1 \\
i_1 & \mbox{ otherwise}
\end{cases}
\end{align}
\end{definition}

For example, consider two time intervals, $i_1 = \left(4/4/2012, UC \right)$ and $i_2 = \left(24/4/2012, UC \right)$. The result of applying the CloseR$\left(i_1, i_2 \right)$ is $i_1 = \left(4/4/2012, 23/4/2012 \right)$.

@@

\begin{definition}
\label{def:close-current-version}
\emph{Close-current$\left(r, t\right)$}.
Consider the elements in definition \ref{def:valid-time-relation}. The function Close-current$\left(r, t \right)$ closes any current version $t_k$ of the entity given by $t$ if it exists and add the new version $t$. For the implementation the following variables are defined:
\begin{align}
 \label{eq:variables-current-version}
t_{CUR} = \mbox{ Current } \left(r, t[A_K] \right)\\
\nonumber
t_{UP} = t_{CUR}; t_{UP}\left[i \right] = \mbox{Close }\left(t_{UP}\left[i \right], t \right)
\nonumber
\end{align}
Then, $t_{CUR}$ is the current version of the entity given by the tuple $t$, and $t_{UP}$ is the updated version of the tuple $t_{CUR}$. In this updated version, the time interval given by $i_{UP}$ is closed with respect to the tuple $t$.

\begin{eqnarray}
\label{eq:close-current}
\text{Close-current} \left(r, t \right) =\\
\begin{cases}
\nonumber
r - t_{CUR} \cup \lbrace t_{UP}\rbrace \cup \lbrace t \rbrace
& \mbox{ if } t_{CUR} \neq \emptyset \\
% R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseR } \left(I, J\right) \right)\right \rbrace \cup \left \lbrace\left(A, J\right) \right \rbrace  \\
% \nonumber
% \mbox{ if  Current } \left(R, A \right) \neq \emptyset\\ %\wedge J > \mbox{ CloseR } \left(I, J \right)   \\
\nonumber r , \text{ otherwise}
\end{cases}
\end{eqnarray}
\end{definition}

For example, consider the document with ID = 3 in Table \ref{tbl:library-sample}. The current version of the document started on 4/4/2012. The document was not valid anymore but, for some reason, the date  was not registered. The document was valid again on 24/4/2012. Then, the Close-current function closes the current version of the document and adds a new version. First, the function CloseR is applied with $i_1 = \left(4/4/2012, UC \right)$ and $i_2 = \left(24/4/2012, UC \right)$. Hence, the value for the time interval $i_1$ is $ \left(4/4/2012, 23/4/2012 \right)$. Then, the modifications on the value for the time interval $i_1$ are stored. Finally a new row with the current values of the document and the time interval $i_2$ is stored.
% patient with ID = 3 in Table \ref{tbl:library-sample}. The current version arrived on 4/4/2012. The patient left the hospital but, for some reason, the date  was not registered. The patient arrived again at the hospital on 24/4/2012. Then, the Close-current function allows to close the current version of the patient and insert a new version. First, the function CloseR is applied with $I = \left(4/4/2012, UC \right)$ and $J = \left(24/4/2012, UC \right)$. Hence, the value for the time interval $I$ is $ \left(4/4/2012, 23/4/2012 \right)$. Then, the modifications on the value for the time interval $I$ are stored. Finally, a new row with the current values of the patient and the time interval $J$ is stored.
The result of this operation is illustrated on Table \ref{tbl:library-sample-2}.

%It is also possible to append at the beggining by using \eqref{def:close-a-crisp-interval-l} and \eqref{def:first-in-relation}.
%\begin{definition}
%\label{def:append-first-version}
%Consider an object $A$, the relation $R$ and a time interval $J$. Then, the function append-first$\left(R,A,J \right)$ closes any right open version of the object $A$ if it exists and add the new version $\left(A, J \right)$.
%
%\begin{eqnarray}
%%\label{eq:close-current}
%\text{Append-first} \left(R, A, J \right) =\\
%\begin{cases}
%\nonumber
%\big \lbrace R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseL } \left(I, J\right) \right) \cup \left(A, J\right)\right \rbrace  \big \rbrace \\
%\nonumber
%\mbox{ if } I = \mbox{ First } \left(R, A \right) \wedge J < \mbox{ CloseL } \left(I, J \right)   \\
%\nonumber R , \text{ in any other case}
%\end{cases}
%\end{eqnarray}
%\end{definition}
\subsubsection{\label{subsubsec:modify}Modify}
This operation adds new information about an existing entity (given by the tuple $t$) in the instance $r$ of the relation $R$. The modify operation does not remove any previous value of the entity. It closes the current version and adds a new version.

@@

\begin{definition}
 \emph{Modify(r, t)}.
Consider the elements in definition \ref{def:valid-time-relation}. The algorithm for the modify operation is defined as follows.
\end{definition}

\begin{align}
\label{eq:modify}
\MOD \left( r, t\right) =\\
%\begin{cases}
\nonumber
\mbox{ Close-current }\left(r, t\right) %\\
% \mbox{ if Current}\left(R ,A \right) \neq \emptyset \\
% R,  \mbox{ otherwise }
%\end{cases}
\end{align}

\subsubsection{\label{subsubsec:insert}Insert}
The user wants to store an entity (given by the tuple $t$) which is valid in the instance $r$ of the relation $R$ during the time interval specified by $i = \left(s, e \right)$.
%
%
%The interpretation of the insert operation is the following: The user wants to store in the database that an object $A$  is true for some period(s) of time $I$ in the given relation $R$. To indicate that the object $A$ is current in the relation, the special value Until Current, $\UC$  is used. 
There are two main cases when performing a create operation:
\begin{enumerate}
\item The entity was never in the relation: The entity is added with the valid-time indicated by the crisp interval $i$. 
% In the example, this correspond with the following insertion sentences:
% \begin{verbatim}
% Insert(3,'Dracula',
% (15/3/2012, 30/3/2012));
% Insert(4,'Frankenstein',
% (25/3/2012, 4/4/2012));
% Insert(5,'Harry Potter',
% (18/3/2012, 2/4/2012));
% \end{verbatim}

\item The entity is in the relation. Depending on the value of the time interval, there are three possibilities:
	\begin{enumerate}
	\item Insert $t$ in the instance $r$ of the relation $R$. If the time interval $i$ does not overlap any other valid-time interval in the instance $r$ relation $R$ for the entity. For example, consider that the document with ID = 3 began again to be valid on 4/4/2012 and it is still currently valid. This is illustrated in Table \ref{tbl:library-sample}.
%  The insert sentence is:
% 	      \begin{verbatim}
% Insert(3,'Dracula',
% (4/4/2012, UC));
% 	      \end{verbatim}

	\item Modify and close the current version of $t$ and insert the new version. For example, consider now that the document with ID = 3 was valid on 24/4/2012. Here the problem is that the document with ID = 3 was valid on 4/4/2012 but, for some reason, the ending date was not stored. If the document is again valid, then it is necessary to set the ending date and add a new row with the new starting date. This is illustrated in Table \ref{tbl:library-sample-2}.
%  the insert sentence is:
% 	    \begin{verbatim}
% Insert(3,'Dracula',
% (24/4/2012, UC));
% 	      \end{verbatim}
	\item Reject the insertion, if the time interval $i$ does overlap any existing valid-time interval for the entity $t$ in the instance $r$ of the relation $R$. For example, consider that the document manager wants to introduce a past valid-time for the document with ID = 3. The validity period for the document is $[$6/4/2012, 25/4/2012$]$. As the dates do overlap, it is not possible that the document was valid at that time interval. Therefore, the insertion is rejected.
%  The insert sentence is:
% 
% 	      \begin{verbatim}
% Insert(3,'Dracula',
% (6/4/12/2012,25/4/2012));
% 	    \end{verbatim}

@@

\end{enumerate}
\begin{samepage}
\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample-2}Sample historical database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Entity  & Start & End \\
\hline
3 & E.U. & 15/3/2012 & 30/3/2012 \\
4 & N.A.T.O. & 25/3/2012 & 4/4/2012 \\
5 & C.E.I.  & 18/3/2012 & 2/4/2012\\
3 & E.U. & 4/4/2012 & 23/4/2012 \\
3 & E.U. & 24/4/2012 & UC \\
\hline\\
\end{tabular}
}
\end{samepage}

\begin{definition}
 \emph{Insert(r, t)}.
Consider the elements in definition \ref{def:valid-time-relation}. Then, the algorithm for the implementation of the insert operation is defined as follows.
\end{definition}
